Algebraic Groups\vspace{-1ex}

### 2019-10-02 Algebraic Geometry/Algebraic-Geometry.tex
Update

\usepackage{amsmath, graphicx, color, fancyhdr, tikz-cd, mdframed, enumitem, framed, adjustbox, bbm, upgreek, xcolor, hyperref, manfnt}
\usepackage[framed,thmmarks]{ntheorem}
\usepackage[style=alphabetic, bibencoding=utf8]{biblatex}
%Set the bibliography file
\bibliography{sources}

@@

>>>>

\usepackage{amsmath, graphicx, color, fancyhdr, tikz-cd, mdframed, enumitem, framed, adjustbox, bbm, upgreek, xcolor, hyperref, manfnt, wasysym}
\usepackage[framed,thmmarks]{ntheorem}
\usepackage[style=alphabetic, bibencoding=utf8]{biblatex}
%Set the bibliography file
\bibliography{sources}

@@

\section{October 2, 2019}
\subsection{Questions without (complete) answers}
\subsubsection{Morphisms and stuff}
A question to get things started for the day. Let $X$ and $Y$ be spaces with functions and let $Y$ be an affine variety and let $f:Y\to X$ be a map of sets 
(but with no further assumption on $f$). This naturally induces amap from $\calO_X(X)$ to the functions $\Hom_{\Set}(Y,k)$ (which 
clearly contains the regular functions on $Y$).

Further assume that there exists a $\gamma:\O_X(X)\to\O_Y(Y)$. We know that since $Y$ is affine, $\gamma$ corresponds to a morphism $\varphi:Y\to X$.
Then the question is: when does $f=\varphi$? We've already answered this question for $\A1$, notice.
\subsubsection{Algebraic closure}
Where do we use algebraic closure of the base field? It has been swept under the rug a bit, but consider the function
\[\frac{1}{x^2+1}:\bbR\to \bbR.\]

This certainly seems like it should be a regular function (e.g. it is rational and defined everywhere on $\bbR$) but this 
conflicts with the idea that we want to identify $\O_{\A1}(\A1)=k[t]$, but that is clearly not the case here. Think about this.
\subsubsection{Yet another}
Consider the set $R$ of all continuous maps $k\to k$ under the cofinite topology. Someone asked if $R$ is a $k$-algebra. The answer is 
a bit convoluted, but the short answer is no. Specifically if we are using the product topology on $k\times k$, the addition map isn't continuous! 
This also points to the question of what topology is the correct one to use on these things.

\subsection{Back to affine varieties}
Recall that we constructed a (highly-non-canonical) picutre of how any affine variety arises as a closed subset of some affine space $k^n$.

We want to remove this dependence on presentation, however, and that is what we are working toward.

\subsubsection{Affine Space}
Now we focus in on $\A n=k^n$. We really want that the projection functions $x_i:k^n\to k$ should be regular. But since we want this (eventually)
to form a $k$-algebra, we want that each $f\in[x_1,\dots,x_n]$ should be regular!

The axioms of a space with functions tells us that the \textbf{vanishing locus}
\[Z(f)=\{a|f(a)=0\}\subseteq k^n\]
and furthermore $Z(S)$ should be closed for all $S\subseteq k[x_1,\dots,x_n]$. This leads us to a definition:
\begin{defn}
	A subset $Z\subseteq k^n$ is \textbf{Zariski-closed} if there exists an $S\subseteq k[x_1,\dots,x_n]$
	such that $Z=Z(S)$.
\end{defn}
\begin{lem}
	The Zariski closed sets are the closed sets of a topology (called the \textbf{Zariski Topology}).
\end{lem}
\begin{prf}
	Just do it. Nike. \checked
\end{prf}
\begin{rmk}
	Notice that here the set $\{(a,-a)\}\subseteq k^2$ (the pullback of zero under the addition map) is Zariski closed!
	This fixes the problem we were running into in the third question above.
\end{rmk}
Now since $Z(S)=Z(I_S)$ where $I_S$ is the ideal generated by $S$, it is enough to consider vanishing loci of ideals. Furthermore we have the 
map that extracts the ideal of functions that vanish on a set $Z\subseteq k^n$. There are a ton of great identiies you can prove here. Go to your 
favorite algebra book (e.g. Dummit \& Foote) to see them.

\subsubsection{Functions}
What about functions on these spaces? If we take $f\in k[x_1,\dots,x_n]$ these seem like they should be regular functions $k^n\to k$.
\begin{thm}[(Weak) Nullstellensatz]
	Say $k=\bar k$. Then every maximal ideal $\frakm\lhd k[x_1,\dots,x_n]$ has the form $(x_1-a_1,\dots,x_n-a_n)$.
\end{thm}
\begin{rmk}
	Equivalently, it is the kernel of a $k$-algebra morphis $k[x_1,\dots,x_n]\to k$.
\end{rmk}
\begin{cor}[Nullstellensatz]
	Let $J$ be an ideal of $k[x_1,\dots,x_n]$. Then $I(Z(J))=\sqrt{J}$.
\end{cor}
\begin{prf}
	\textit{Notice this only works when $k$ is uncountable!} Suppose that $\frakm$ is a maximal ideal with residue field 
	$L=k[x_i]/\frakm$. This gives us a surjection of $k[x_1,\dots,x_n]\to L$. Thus $\dim_k L$ is countable!

	But $\dim_k k(t)$ is uncountable! The proof here is that the $\frac{1}{t-\lambda}$ for $\lambda\in k$ is a linearly-independent collection.
	So then $L/k$ is algebraic, and since $k=\bar k$ $L=k$.
\end{prf}

### 2019-10-02 Algebraic Groups/Algebraic-Groups.tex
Update

>>>>

\section{October 2, 2019}
Notice that a \textbf{relative group scheme} (referred to in last lecture) refers to a groups scheme over an arbitrary 
base scheme $S$.

\subsection{Properties of schemes}
Today we are going to be talking about reducedness, connectedness, irreduciblility, regularity, and smoothness.

Recall that a scheme $X$ is \textbf{reduced} if and only if $\forall x\in X$, $\calO_{X,x}$ is reduced. An example of a non-reduced 
scheme is $\Spec k[x]/(x^2)$.
\begin{defn}
	We say a scheme $X$ over $k$ is \textbf{geometrically reduced} if for all field extensions $k'/k$,
	\[X_{k'}=X\times_{\Spec k}\Spec k'\]
	is reduced.
\end{defn}
\begin{rmk}
	It is equivalent that $X_{\bar k}$ is reduced if and only if every $k'/k$ is purely inseperable (I think).
\end{rmk}
\begin{rmk}
	If $k$ is perfect, then $X$ is reduced if and only if $X$ is geometrically reduced.
\end{rmk}
\begin{defn}
	A local ring $(A,\frakm)$ is \textbf{regular} if $\dim _{\text{Krull}}A=\dim_{A/\frakm} \frakm/\frakm^2$
\end{defn}
\begin{defn}
	A scheme $X$ is regular if for all $x\in X$, $\calO_{X,x}$ is regular.
\end{defn}
\begin{rmk}
	If $X\to \Spec k$ and $x\in X(k)$, the tangent space at $x$ is 
	\[T_{X,x}=(\frakm/\frakm^2)^\vee=\{f:\Spec k[\varepsilon]/\varepsilon^2\to X|0\mapsto x\}\]
\end{rmk}
\begin{rmk}
	Notice that if $X\to \Spec k$ is regular and $k'/k$ is a field extension, then $X_{k'}$ is not necessarily regular.
\end{rmk}

\begin{defn} 
	A Scheme $X\to \Spec k$ of finite type is \textbf{smooth} if $X_{\bar k}$ is regular.
\end{defn}

\subsection{Facts about algebraic groups}
Then we can return to the proposition we want to prove:
\begin{prop}
	Let $G\to \Spec k$ be an algebraic group. Then $G$ is geometrically reduced if and only if $G$ is smooth over $\Spec k$.
\end{prop}
\begin{prf}
	Smoothness over $k$ implies reducedness. Now since we are only interested in the algebraic closure of $k$, we can say $k=\bar k$. Because $G$ 
	is reduced, there exists a nonempty open $U\subseteq G$ that is smooth. Then since $G(k)\subseteq |G|$ is dense in $G$ (as a topological space) 
	and Then $G=\cup_{g\in G(k)}m_g(U)$ for our smooth $U$, and this gives us a smooth cover of $G$.
\end{prf}

We will see next itme that all linear algebraic groups over $k$ where $\ch k=0$ are all 
geometrically reduced (and thus smooth).

\subsection{Connectedness}
Let $G$ be an algebraic group over $k$. Then we have our maps $e:\Spec k\to G$, so consider it as $e\in G(k)$.
Let $G^0\subseteq G$ be the connected component of $e$. It is both open and closed.
\begin{rmk}
	If $X\to \Spec k$ is of finite type and $x\in X(k)$, then $X$ being connected implies that $X$ is geometrically connected.
\end{rmk}
This establishes that $G^0$ is actually geometrically connected! We actually will see
\begin{prop}
	$G^0\subseteq G$ is an (open and closed) algebraic subgroup.
\end{prop}
The idea here is that $G^0\times G^0$ is connected, so the image of the multipication map on this set lands in a connected component (since it is connected).
Since $e\in G^0$, and $m(e,e)=e\in G^0$, this shows that the multiplication map restricts to a well-defined map $G^0\times G^0\to G^0$. A similar argument 
goes through for the inverset map, etc.

The upshot here is that if $G$ is an algebraic group, then there exists a factorization 
\[1\to G^0\to G\to \pi_0(G)\to 1\]
where $\pi_0(G)$ is given the structure of a discrete group.
\begin{rmk}
	Now we also have that $(G^0)_{k'}=(G_{k'})^0$ for all $k'/k$. The idea is to get a map of of one into the other and then use clopenness and connectedness 
	to show they are equal.
\end{rmk}

\begin{prop}
	A connected algebraic group over $k$ is irreducible.
\end{prop}
\begin{prf}
	We can assume $k=\bar k$. Suppose $G= X\cup Y$, where both are closed, $X$ is irreducible, 
	and $X\cap Y\ne\varnothing$. Thus there exists an element $g\in X\setminus Y$. That is, $g$ lies in a single irreducible component.

	But then using the multiplication by $h$ map on $G$, we get to every other point in $G$, so every point 
	is in a single irreducible component. But the intersection was nontrivial! Or something.
\end{prf}

\begin{prop}
	If $G_{\text{red}}$ is geometrically connected, then $G_{\text{red}}\subseteq G$ is a subgroup. In particular, 
	if $k$ is perfect, then $G_{\text{red}}$ is a subgroup of $G$.
\end{prop}
\begin{rmk}
	$X$ is geometrically reduced implies that $X\times X$ is geometrically reduced.
\end{rmk}
